Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-06-01 manuscript.tex
tbl lab

\noindent\textbf{Public significance statement}: In this paper, we provide evidence that the fulfillment of situational needs during real-life intergroup contacts meaningfully predicts perceived interaction quality and positive outgroup attitudes. Methodologically, this offers testament to the emerging practice of capturing real-life interactions using extensive longitudinal data. Theoretically, our results give weight to motivational fulfillment as a flexible and effective mechanism for understanding positive intergroup contact.
}

@@

\section{The Present Research}
Using three independent sets of extensive longitudinal data (Studies 1–3), the aim of this paper is essentially threefold. We (1) seek to test the basic ideas of the contact theory within real-world experience sampling data. We (2) aim to test the core need fulfillment mechanism within the real-world data. And we (3) seek to ensure the stability, robustness, and embeddedness of our results.

Firstly, for the general contact hypothesis test, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life extensive longitudinal data. Translating the contact hypothesis into extensive longitudinal data is not a trivial task, as past research traditions have used two fundamentally different approaches. While lab studies have tended to focus on the effect of a single positive interaction, cross-section studies have primarily investigated the frequency of positive interactions more generally. Extensive longitudinal data allows us to investigate both. We can test whether having a specific type of interaction vs. not having an interaction improves intergroup relations, but we can also use the participant's 30-day contact reports to test whether participants with more positive interactions tend to benefit more from intergroup contact. Testing both approaches to the contact hypothesis allows us to go beyond a replication of the basic theory but could disentangle individual- from aggregated contact effects and would allow for a direct comparison with both bodies of literature.

We test the basic contact hypothesis within and across the three studies. In particular, we assess the effect of individual interactions within each study using a multilevel model, but to avoid power limitations, we test the collective effect of contact frequency and -quality after the individual studies, across all participants.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H1:] Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for more favorable outgroup attitudes within and across extensive longitudinal data.
\end{enumerate}

The test of Allport's conditions is notably restricted to measurements that report on outgroup interactions because Allport's conditions and interaction quality ratings cannot meaningfully be measured or imputed if participants did not have an interaction. Focusing on the interactions in detail, we use a multilevel regression model to test whether interactions that are higher in the fulfillment of Allport's conditions predict more favorable outgroup attitudes. We would also expect that such interactions are perceived as higher in interaction quality.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H2:] Based on the literature about Allport’s optimal contact conditions, intergroup interactions that are higher in equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data.
\end{enumerate}

Once the general contact hypothesis is established within the ESM data, our second main aim is to test our main theoretical proposal that the fulfillment of situational needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. As our main proposal is concerned with the mechanisms of successful intergroup contact, we again focus on outgroup interaction reports. Within a multilevel model, we expect interactions that are higher in situational core need fulfillment to be perceived as more positive, and as a result that these interactions also predict more positive outgroup attitudes. We also expect the needs mechanism to work at least as well as Allport's conditions. We particularly expect part of Allport's contact conditions to be a static set of situational needs so that the situational core need fulfillment should explain some of the same variance in outgroup attitudes.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H3:] Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the extensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
\end{enumerate}

@@

A second non-trivial aspect of translating the intergroup contact hypothesis into extensive real-world data was the choice of the outcome variable. For our main analyses, we chose outgroup attitudes --- the positive or negative evaluation of the other group. We chose outgroup attitudes mainly because they are the most common outcome considered within the intergroup contact literature \citep[e.g.,][]{Pettigrew2006, Paolini2021}. As the methodology is relatively new to the field, we sought to first replicate (and then extend) the most reliable effects of the contact hypothesis within the ESM data. Outgroup attitudes are, however, not without controversy, especially for minority group members. Positive outgroup attitudes can increase harmony and reduce the willingness to support social change among the disadvantaged in some cases --- even in the face of injustice \citep[e.g.,][]{dixon2012, saguy2009}. While a recent review found the effect to be less conclusive for longitudinally collected data and less consistent for positive interactions (rather than interactions generally), the backfire effect remains an important possibility for the present data \citep[see][]{reimer2023}. In order to ensure at least a direct benefit to the minority group members, we also assess the effect of the need fulfillment mechanism on well-being as the dependent variable as part of our robustness analyses below.

@@

\section{Discussion}
% aims re-iterated
The main aim of this project was to test the basic tenets of the intergroup contact hypothesis and Allport's optimal conditions in real-life extensive longitudinal data as well as to test whether the fulfillment of situational needs meaningfully predicts positive interaction perceptions and outgroup attitudes.

@@

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results regarding the importance of situational key need fulfillment, we find that in all three extensive longitudinal data sets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions, more positive outgroup attitudes, as well as higher well-being. We also find that in all three studies, need fulfillment and perceived interaction quality likely shared a large part of the variance they explained in outgroup attitudes (when considering partial regression coefficients in a joint model). We would like to reiterate here that we specifically did not seek to test a mediation-style model. Even though the shared explained variance was predicted in our pre-registration based on a theoretical model, and the pattern was stable across the three studies, it is important to note that the data is none-causal, and the effect might alternatively be driven by an unobserved third variable or by multicollinearity.

@@

A second limitation lies in our methodology. While extensive longitudinal data is close to real-life events, this method comes at the expense of longer and more robust scales. Long and repetitive scales are often not feasible in extensive longitudinal methods because of the increased burden to the participants. To circumvent this shortcoming, we have ensured that the measures we used were, whenever possible, based on past validations. However, the circumstance remains that extensive longitudinal data often does not allow the same scrutiny of measurement reliability as single–shot cross-sectional data sets. An additional methodological question lies in the unexplored potential of the longitudinal aspects of our data. For our research questions, we have focused on contemporaneous effects within the data set, yet future investigations should seek to extend the mechanism to developmental trajectories within and between participants.

@@

\subsection{Implications}
% ESM data
Despite these limitations, we can nonetheless draw a number of implications for other researchers and practitioners --- ranging from the benefits of longitudinal data to theoretical implications. A first implication concerns the feasibility and usefulness of extensive longitudinal data for intergroup contact research and the broader field of social psychology. While setting up an extensive longitudinal study is not easy, we believe the efforts to be similar to a sizable cross-sectional data collection (i.e., for a longitudinal or a high-quality cross-sectional data set with over 3,000 intergroup interactions captured and over 10,000 data points in total). Intensive longitudinal data, thus, opens up the possibility to explore research questions that focus on real-life phenomena outside the lab or focus on phenomena that depend on changes and influences over time. In the context of intergroup contact research, we are among the first to answer calls to test intergroup contact mechanisms using extended real-life data \citep[e.g.,][]{Pettigrew2011, MacInnis2015}. In doing so, we not only collected an unprecedented amount of real-life data, but our consideration of extensive longitudinal data may present new inconsistencies in how participants perceive and cognitively aggregate their past interactions with other groups — which may suggest large-scale recall biases or conflations of within and between participant effects in conventional cross-section studies.

@@

\subsection{Conclusion}
% Conclusion paragraph
In sum, we used extensive longitudinal methodologies to capture real-life interactions of recent migrants with the majority outgroup. Our three studies showcase the feasibility and utility of such data to test intergroup contact theory. We provide evidence that the fulfillment of situational needs during real-life intergroup contacts meaningfully predicts perceived interaction quality and positive outgroup attitudes. Our results point to motivational needs as an understudied aspect of intergroup contact that is important in understanding when and why an interaction is perceived as positive and will lead to more positive outgroup attitudes.

@@

\subsection{1. Contact hypothesis in extensive longitudinal data.} 
Within the individual studies, we begin by testing the most basic assumption of the intergroup contact hypothesis that outgroup attitudes should be more positive after outgroup interactions but not after non-outgroup interactions. For this we use multilevel regression analyses, predicting outgroup attitudes from outgroup interaction and non-outgroup interaction dummy variables. We also include the participant means as level two predictors to fully disentangle within- (level 1) and between-participant (level 2) effects of intergroup contact \citep[e.g.,][Section 4.6]{snijders2012}.

\begin{mdframed}[style=mdfhypothesis]
    \begin{hyp}[H\ref{hyp:contactHyp}] \label{hyp:contactHyp}
        Based on the most general understanding of the contact hypothesis, positive intergroup contacts should be associated with more favorable outgroup attitudes across extensive longitudinal data.
    \end{hyp}

@@

\subsection{3. Allport's conditions in extensive longitudinal data.}
Within the third study, we formally measure all of Allport's optimal contact conditions. We use multilevel regression models to test whether the fulfillment of Allport's conditions in real-life data predicts more positive outgroup attitudes and higher perceived interaction quality, using the same approach as for the core need fulfillment above.

>>>>

\noindent\textbf{Public significance statement}: In this paper, we provide evidence that the fulfillment of situational needs during real-life intergroup contacts meaningfully predicts perceived interaction quality and positive outgroup attitudes. Methodologically, this offers testament to the emerging practice of capturing real-life interactions using intensive longitudinal data. Theoretically, our results give weight to motivational fulfillment as a flexible and effective mechanism for understanding positive intergroup contact.
}

@@

\section{The Present Research}
Using three independent sets of intensive longitudinal data (Studies 1–3), the aim of this paper is essentially threefold. We (1) seek to test the basic ideas of the contact theory within real-world experience sampling data. We (2) aim to test the core need fulfillment mechanism within the real-world data. And we (3) seek to ensure the stability, robustness, and embeddedness of our results.

Firstly, for the general contact hypothesis test, our study is among the first to test the fundamental tenets of intergroup contact and Allport's conditions in real-life intensive longitudinal data. Translating the contact hypothesis into intensive longitudinal data is not a trivial task, as past research traditions have used two fundamentally different approaches. While lab studies have tended to focus on the effect of a single positive interaction, cross-section studies have primarily investigated the frequency of positive interactions more generally. Intensive longitudinal data allows us to investigate both. We can test whether having a specific type of interaction vs. not having an interaction improves intergroup relations, but we can also use the participant's 30-day contact reports to test whether participants with more positive interactions tend to benefit more from intergroup contact. Testing both approaches to the contact hypothesis allows us to go beyond a replication of the basic theory but could disentangle individual- from aggregated contact effects and would allow for a direct comparison with both bodies of literature.

We test the basic contact hypothesis within and across the three studies. In particular, we assess the effect of individual interactions within each study using a multilevel model, but to avoid power limitations, we test the collective effect of contact frequency and -quality after the individual studies, across all participants.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H1:] Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for more favorable outgroup attitudes within and across intensive longitudinal data.
\end{enumerate}

The test of Allport's conditions is notably restricted to measurements that report on outgroup interactions because Allport's conditions and interaction quality ratings cannot meaningfully be measured or imputed if participants did not have an interaction. Focusing on the interactions in detail, we use a multilevel regression model to test whether interactions that are higher in the fulfillment of Allport's conditions predict more favorable outgroup attitudes. We would also expect that such interactions are perceived as higher in interaction quality.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H2:] Based on the literature about Allport’s optimal contact conditions, intergroup interactions that are higher in equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the intensive longitudinal data.
\end{enumerate}

Once the general contact hypothesis is established within the ESM data, our second main aim is to test our main theoretical proposal that the fulfillment of situational needs is meaningfully related to more positive outgroup attitudes following intergroup interactions. As our main proposal is concerned with the mechanisms of successful intergroup contact, we again focus on outgroup interaction reports. Within a multilevel model, we expect interactions that are higher in situational core need fulfillment to be perceived as more positive, and as a result that these interactions also predict more positive outgroup attitudes. We also expect the needs mechanism to work at least as well as Allport's conditions. We particularly expect part of Allport's contact conditions to be a static set of situational needs so that the situational core need fulfillment should explain some of the same variance in outgroup attitudes.
\begin{enumerate}[leftmargin=1.5\parindent]
    \item[H3:] Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions within the intensive longitudinal data. We also expect situational need fulfillments to work at least as well as Allport's optimal contact conditions in predicting outgroup attitudes.
\end{enumerate}

@@

A second non-trivial aspect of translating the intergroup contact hypothesis into intensive real-world data was the choice of the outcome variable. For our main analyses, we chose outgroup attitudes --- the positive or negative evaluation of the other group. We chose outgroup attitudes mainly because they are the most common outcome considered within the intergroup contact literature \citep[e.g.,][]{Pettigrew2006, Paolini2021}. As the methodology is relatively new to the field, we sought to first replicate (and then extend) the most reliable effects of the contact hypothesis within the ESM data. Outgroup attitudes are, however, not without controversy, especially for minority group members. Positive outgroup attitudes can increase harmony and reduce the willingness to support social change among the disadvantaged in some cases --- even in the face of injustice \citep[e.g.,][]{dixon2012, saguy2009}. While a recent review found the effect to be less conclusive for longitudinally collected data and less consistent for positive interactions (rather than interactions generally), the backfire effect remains an important possibility for the present data \citep[see][]{reimer2023}. In order to ensure at least a direct benefit to the minority group members, we also assess the effect of the need fulfillment mechanism on well-being as the dependent variable as part of our robustness analyses below.

@@

\section{Discussion}
% aims re-iterated
The main aim of this project was to test the basic tenets of the intergroup contact hypothesis and Allport's optimal conditions in real-life intensive longitudinal data as well as to test whether the fulfillment of situational needs meaningfully predicts positive interaction perceptions and outgroup attitudes.

@@

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results regarding the importance of situational key need fulfillment, we find that in all three intensive longitudinal data sets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions, more positive outgroup attitudes, as well as higher well-being. We also find that in all three studies, need fulfillment and perceived interaction quality likely shared a large part of the variance they explained in outgroup attitudes (when considering partial regression coefficients in a joint model). We would like to reiterate here that we specifically did not seek to test a mediation-style model. Even though the shared explained variance was predicted in our pre-registration based on a theoretical model, and the pattern was stable across the three studies, it is important to note that the data is none-causal, and the effect might alternatively be driven by an unobserved third variable or by multicollinearity.

@@

A second limitation lies in our methodology. While intensive longitudinal data is close to real-life events, this method comes at the expense of longer and more robust scales. Long and repetitive scales are often not feasible in intensive longitudinal methods because of the increased burden to the participants. To circumvent this shortcoming, we have ensured that the measures we used were, whenever possible, based on past validations. However, the circumstance remains that intensive longitudinal data often does not allow the same scrutiny of measurement reliability as single–shot cross-sectional data sets. An additional methodological question lies in the unexplored potential of the longitudinal aspects of our data. For our research questions, we have focused on contemporaneous effects within the data set, yet future investigations should seek to extend the mechanism to developmental trajectories within and between participants.

@@

\subsection{Implications}
% ESM data
Despite these limitations, we can nonetheless draw a number of implications for other researchers and practitioners --- ranging from the benefits of longitudinal data to theoretical implications. A first implication concerns the feasibility and usefulness of intensive longitudinal data for intergroup contact research and the broader field of social psychology. While setting up an intensive longitudinal study is not easy, we believe the efforts to be similar to a sizable cross-sectional data collection (i.e., for a longitudinal or a high-quality cross-sectional data set with over 3,000 intergroup interactions captured and over 10,000 data points in total). Intensive longitudinal data, thus, opens up the possibility to explore research questions that focus on real-life phenomena outside the lab or focus on phenomena that depend on changes and influences over time. In the context of intergroup contact research, we are among the first to answer calls to test intergroup contact mechanisms using extended real-life data \citep[e.g.,][]{Pettigrew2011, MacInnis2015}. In doing so, we not only collected an unprecedented amount of real-life data, but our consideration of intensive longitudinal data may present new inconsistencies in how participants perceive and cognitively aggregate their past interactions with other groups — which may suggest large-scale recall biases or conflations of within and between participant effects in conventional cross-section studies.

@@

\subsection{Conclusion}
% Conclusion paragraph
In sum, we used intensive longitudinal methodologies to capture real-life interactions of recent migrants with the majority outgroup. Our three studies showcase the feasibility and utility of such data to test intergroup contact theory. We provide evidence that the fulfillment of situational needs during real-life intergroup contacts meaningfully predicts perceived interaction quality and positive outgroup attitudes. Our results point to motivational needs as an understudied aspect of intergroup contact that is important in understanding when and why an interaction is perceived as positive and will lead to more positive outgroup attitudes.

@@

\subsection{1. Contact hypothesis in intensive longitudinal data.} 
Within the individual studies, we begin by testing the most basic assumption of the intergroup contact hypothesis that outgroup attitudes should be more positive after outgroup interactions but not after non-outgroup interactions. For this we use multilevel regression analyses, predicting outgroup attitudes from outgroup interaction and non-outgroup interaction dummy variables. We also include the participant means as level two predictors to fully disentangle within- (level 1) and between-participant (level 2) effects of intergroup contact \citep[e.g.,][Section 4.6]{snijders2012}.

\begin{mdframed}[style=mdfhypothesis]
    \begin{hyp}[H\ref{hyp:contactHyp}] \label{hyp:contactHyp}
        Based on the most general understanding of the contact hypothesis, positive intergroup contacts should be associated with more favorable outgroup attitudes across intensive longitudinal data.
    \end{hyp}

@@

\subsection{3. Allport's conditions in intensive longitudinal data.}
Within the third study, we formally measure all of Allport's optimal contact conditions. We use multilevel regression models to test whether the fulfillment of Allport's conditions in real-life data predicts more positive outgroup attitudes and higher perceived interaction quality, using the same approach as for the core need fulfillment above.
